---
title: "Explore term 1 & term2 exam data"
output:
  pdf_document: default
  html_document:
    keep_md: yes
  word_document: default
---
========================================================  
## Introduction
Exploring exam data of term one and two.

## Preparation
### Setting up working environment

* Working Directory  
A directory named 'Exam' is built as the main working directory of this work 

* Software Packages
To finish the project, a couple of software needs to be installed.   
-- R Version 4.1.0 64bit  
-- RStudio Version 1.4.1717  
-- notepad++ Version 7.9.1  
The following packages in R are expected to be used:  
-- knitr - markdown tools  
-- ggplot2  
-- dplyr 
-- tidyr 
-- Kendall 
-- nortest 
-- effectsize 
-- HH 

### Loading and preprocessing the data

Load the data, which can be found at 
[https://github.com/sunbaoshi1975/UWThesisDataset](https://github.com/sunbaoshi1975/UWThesisDataset). 

The original data is assumed to be placed in the working directory.

```{r loadLibs, echo=FALSE, results='hide', include=FALSE}
# Notes: if fail to load library, use install.packages("Package Name") to download and install it first.
#install.packages("tidyr")
#install.packages("effectsize")
#install.packages('HH')

library("dplyr")
library(tidyr)
library(ggplot2)
library(scales)
library(Kendall)
library(nortest)
library(methods)
library(effectsize)
library(HH)
```

```{r commonDef}
# some constants
strOperationLog <- "operation_log.csv"
strDataFileList <- "examDataFileList.csv"

## This path should be changed accordingly when performing reproducible research on your computer
#setwd("D:\\Exam")

```

```{r commonFunctions}
colApply <- function(dat, cols = colnames(dat), func = as.factor) {
  dat[cols] <- lapply(dat[cols], func)
  return(dat)
}

loadRawData <- function(filename, exGroup, exRoom, exTerm)
{
  if (file.exists(filename)) {
    suppressWarnings(raw.data <- read.csv(filename, header=TRUE, na.strings=c("NA","#DIV/0!","")))

    ## Filter data
    columns <- c("RID")
    columns <- c(columns, "Gender")
    columns <- c(columns, "Chinese")
    columns <- c(columns, "Math")
    loaded <- raw.data[!is.na(raw.data$RID), columns]

    ## Expand columns
    ### Total, Group, Room, Term
    loaded <- cbind(loaded, Total=loaded$Chinese+loaded$Math, Group=exGroup, Room=exRoom, Term=exTerm)
  }
}

meanRound <- function(x, digits = 1, trim = 0, na.rm = FALSE, ...) {
  return(round(mean(x, trim=trim, na.rn=na.rm, ...), digits=digits))
}

## Calculate changes between the two terms
getChangedColumn <- function(x, name, num) {
  changes = c()
  for (i in 1:num) {
    changes <- c(changes, NA, x[name][i*2,1]-x[name][i*2-1,1])
  }
  return(changes)
}

## Calculate increments of experimental group, control group and their gap
computeIncrement <- function(x, y, g) {
  incExp <- mean(x[g=='Experimental']) / mean(y[g=='Experimental'])
  incControl <- mean(x[g=='Control']) / mean(y[g=='Control'])
  IncGap <- incExp - incControl
  return(c(incExp, incControl, IncGap))
}
```

```{r loadOperationLog, cache=TRUE}
## Load Operation Log
if (!file.exists(strOperationLog)) {
  ## warn the missing file
	sprintf("File %s not found!", strOperationLog)
} else {
  work.opLog <- read.csv(strOperationLog, header=TRUE, na.strings=c("NA","#DIV/0!",""))
  ## Convert Group, Room, Type, Method to factor
  work.opLog <- colApply(work.opLog, c('Group','Room','Type','Method'))
}
```

```{r loadDataFileList, cache=TRUE}
## Load file list
if (!file.exists(strDataFileList)) {
  ## warn the missing file
	sprintf("File %s not found!", strDataFileList)
} else {
  work.flist <- read.csv(strDataFileList, header=TRUE, na.strings=c("NA","#DIV/0!",""))
}
```

```{r loadMainDataFiles}
for (i in 1:nrow(work.flist)) {
  temp.examData <- loadRawData(work.flist$filename[i], work.flist$group[i], work.flist$room[i], work.flist$term[i])
  if(i == 1) {
    work.df <- temp.examData
  } else {
    work.df <- rbind(work.df, temp.examData)
  }
}

## Convert Gender, Group, Room, Term to factor
work.df <- colApply(work.df, c('Gender','Group','Room','Term'))
## Description of the main date frame
str(work.df)
head(work.df)
tail(work.df)
summary(work.df)
```
It was presumed that there was no significant difference in terms of the initial performance between the control group and the experimental group. Since the lighting mode for both groups was the same by the end of term 1, the scores of term 1 were considered as the initial performance. A two-sample two-tailed t-test was performed and reported no significance on each subject.

Let's check it. 

```{r priori_tTest, echo=TRUE}

data = work.df[work.df$Term=='1', ]

# t-test on Term 1 Chinese
t.test(data$Chinese~data$Group)

# t-test on Term 1 Math
t.test(data$Math~data$Group)

```

## Exploratory Data Analysis

Calculate the mean values of overall performance, by-subject performance, by-gender performance, as well as by-subject-and-gender performance.

Calculate and report the mean values.

```{r exploratoryDataAnalysis}
columns <- c("Chinese")
columns <- c(columns, "Math")
columns <- c(columns, "Total")

# By subject
eda.subject <- aggregate(work.df[, columns], by=list(Term=work.df$Term, Group=work.df$Group), FUN = meanRound)

# By gender
eda.gender <- aggregate(work.df[, columns], by=list(Term=work.df$Term, Gender=work.df$Gender, Group=work.df$Group), FUN = meanRound)

# By classroom
eda.room <- aggregate(work.df[, columns], by=list(Term=work.df$Term, Group=work.df$Group, Room=work.df$Room), FUN = meanRound)

# View the sorted by-classroom list
View(eda.room[order(eda.room$Term, -eda.room$Total),])
```


```{r edaCalulateChangesBySubject}
changes <- getChangedColumn(eda.subject, 'Chinese', 2)
eda.subjectEx <- cbind(eda.subject, chgCN=changes)

changes <- getChangedColumn(eda.subject, 'Math', 2)
eda.subjectEx <- cbind(eda.subjectEx, chgMath=changes)

changes <- getChangedColumn(eda.subject, 'Total', 2)
eda.subjectEx <- cbind(eda.subjectEx, chgTotal=changes)

eda.subjectEx[, c('Term','Group','Chinese','chgCN','Math','chgMath','Total','chgTotal')]

```

```{r edaCalulateChangesByGender}
## Calculate changes between the two terms
changes <- getChangedColumn(eda.gender, 'Chinese', 4)
eda.genderEx <- cbind(eda.gender, chgCN=changes)

changes <- getChangedColumn(eda.gender, 'Math', 4)
eda.genderEx <- cbind(eda.genderEx, chgMath=changes)

changes <- getChangedColumn(eda.gender, 'Total', 4)
eda.genderEx <- cbind(eda.genderEx, chgTotal=changes)

eda.genderEx[, c('Term','Gender','Group','Chinese','chgCN','Math','chgMath','Total','chgTotal')]

```

```{r edaCalulateChangesByClassroom}
# Calculate score changes from Term 2 to term 1 for each classroom
columns <- c("Group")
columns <- c(columns, "Room")
columns <- c(columns, "Chinese")
columns <- c(columns, "Math")
columns <- c(columns, "Total")

# Split & Merge data sets
temp.x=eda.room[eda.room$Term==2,columns]
temp.y=eda.room[eda.room$Term==1,columns]
temp.merged <- merge(temp.x, temp.y, by=c('Group','Room'))
# Another method
#temp.merged <- spread(eda.room[,c('Term','Group','Room','Total')], Term, Total)
# Compute changes
eda.roomEx <- mutate(temp.merged, chgCN=Chinese.x-Chinese.y, chgMath=Math.x-Math.y, chgTotal=Total.x-Total.y)

View(eda.roomEx[,c('Group','Room','chgCN','chgMath','chgTotal')])
```


```{r edaPlot_Overall}
#--------------------------------
## Save as png
png(filename='plotOverall.png')

## Colors
myColors=c('Total'="#3591d1", 'Chinese'="#f04546", 'Math'="#62c76b")
## Draw Plot: multiple-line chart
p <- ggplot(data=eda.subject, aes(x=Term, shape=Group, fill=Group)) + 
  geom_line(aes(y=Total, group=Group, colour="Total")) + 
  geom_point(position=position_dodge(0.1), size=3, aes(y=Total)) +
  geom_text(aes(y=Total,label=round(Total,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  geom_line(aes(y=Chinese, group=Group, colour="Chinese")) + 
  geom_point(position=position_dodge(0.1), size=3, aes(y=Chinese)) +
  geom_text(aes(y=Chinese,label=round(Chinese,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  geom_line(aes(y=Math, group=Group, colour="Math")) + 
  geom_point(position=position_dodge(0.1), size=3, aes(y=Math)) +
  geom_text(aes(y=Math,label=round(Math,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  scale_colour_manual(name="Subject", values=myColors) +
  labs(title="Performance Changes", subtitle="Overvall & By-subject") + xlab("Term") + ylab("Exam Score")

## Show the graph
print(p)
#--------------------------------
```

```{r edaPlot_ByClassroom}
#--------------------------------
## Save as png
png(filename='plotByClassroom.png')

meanChgExp = mean(eda.roomEx[which(eda.roomEx$Group=='Experimental'), 'chgTotal'])
meanChgControl = mean(eda.roomEx[which(eda.roomEx$Group=='Control'), 'chgTotal'])

## Colors
myColorGroup=c('Experimental'="#548687", 'Control'="#8FBC94")

## Draw Plot: histogram
p <- ggplot(data=eda.roomEx, aes(x=Room, y=chgTotal, fill=Group)) + 
  geom_bar(stat='identity') + 
  geom_text(aes(y=chgTotal, label=round(chgTotal,digits=1))) +
  geom_hline(aes(yintercept=meanChgExp,col='Experimental'),linetype=5) +
  geom_hline(aes(yintercept=meanChgControl,col='Control'),linetype=3) +
  scale_fill_manual(name="Group", values=myColorGroup) + 
  scale_colour_manual(name="Mean", values=myColorGroup) +
  labs(title="Performance Changes", subtitle="By classroom") + xlab("Classroom") + ylab("Exam Score")

## Show the graph
print(p)

## Close device
#dev.off()
#--------------------------------
```

```{r edaPlot_ByGender}
#--------------------------------
## Save as png
png(filename='plotByGender.png')

## Colors
myColors2=c('F'="#E69F00", 'M'="#56B4E9")
## Draw Plot: multiple-line chart
p <- ggplot(data=eda.gender, aes(x=Term, y=Total, group=Group, shape=Group, colour=Gender)) + 
  geom_line() + 
  geom_point(position=position_dodge(0.1), size=3) +
  geom_text(aes(y=Total,label=round(Total,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  facet_wrap(facet = ~ Gender) +
  scale_colour_manual(name="Gender", values=myColors2) +
  labs(title="Performance Changes", subtitle="By-gender Overall") + xlab("Term") + ylab("Exam Score")

## Show the graph
print(p)

## Close device
#dev.off()

#--------------------------------
```
```{r edaPlot_ByGenderSubject}
#--------------------------------
## Save as png
png(filename='plotByGenderSubject.png')

## Colors
myColors3=c('Chinese'="#f04546", 'Math'="#62c76b")
## Draw Plot: multiple-line chart
p <- ggplot(data=eda.gender, aes(x=Term, shape=Group)) + 
  geom_line(aes(y=Chinese, group=Group, colour="Chinese")) + 
  geom_point(position=position_dodge(0.1), size=3, aes(y=Chinese)) +
  geom_text(aes(y=Chinese,label=round(Chinese,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  geom_line(aes(y=Math, group=Group, colour="Math")) + 
  geom_point(position=position_dodge(0.1), size=3, aes(y=Math)) +
  geom_text(aes(y=Math,label=round(Math,digits=1)), position=position_dodge(width=0.5), vjust=-0.5) +
  facet_wrap(facet = ~ Gender) +
  scale_colour_manual(name="Subject", values=myColors3) +
  labs(title="Performance Changes", subtitle="By gender and subject") + xlab("Term") + ylab("Exam Score")

## Show the graph
print(p)

## Close device
#dev.off()

#--------------------------------
```

```{r normalityTest, echo=TRUE}
hist(work.df$Total, breaks=20, main="Overall Score Distribution", col="blue", xlab="Chinese + Mathematics")
ad.test(work.df$Total)
```
```{r homoscedasticityTest, echo=TRUE}
## bartlett.test require normality, so we turn to Fligner-Killeen test, which does not require normality
#bartlett.test(work.df$Total~work.df$Group)
fligner.test(work.df$Total~work.df$Group)
```

```{r removalUnpairedRecord}
temp.x=work.df[work.df$Term==2,]
temp.y=work.df[work.df$Term==1,]
temp.merged <- merge(temp.x, temp.y, by=c('Group','Room','RID'))
temp.unpaired <- temp.merged[which(temp.merged$Chinese.x==0 | temp.merged$Math.x==0 | temp.merged$Chinese.y==0 | temp.merged$Math.y==0),]

work.df <- mutate(work.df, Room_RID=paste(Room, RID, sep='_'))
temp.unpaired <- mutate(temp.unpaired, Room_RID=paste(Room, RID, sep='_'))
work.dfPaired <- anti_join(work.df, temp.unpaired, by='Room_RID')
```

```{r calculateChanges}
temp.x=work.dfPaired[work.dfPaired$Term==2,]
temp.y=work.dfPaired[work.dfPaired$Term==1,]
temp.merged <- merge(temp.x, temp.y, by=c('Group','Room','RID','Room_RID','Gender'))
work.dfChanges <- mutate(temp.merged, chgTotal=Total.x-Total.y, chgChinese=Chinese.x-Chinese.y, chgMath=Math.x-Math.y)
```

```{r tTest_OpLog}
columns <- c("Times")

# Manual or Auto
work.opLogType <- aggregate(work.opLog[, columns], by=list(Group=work.opLog$Group, Room=work.opLog$Room, Type=work.opLog$Type), FUN = sum)
work.opLogTypeDaily <- mutate(work.opLogType, timesPerDay=x/662)

# t-test
data=work.opLogTypeDaily[work.opLogTypeDaily$Type=='Manual', ]
t.test(data$timesPerDay~data$Group, alternative="greater")

mean(data[data$Group=='Control', c("timesPerDay")]);sd(data[data$Group=='Control', c("timesPerDay")])
mean(data[data$Group=='Experimental', c("timesPerDay")]);sd(data[data$Group=='Experimental', c("timesPerDay")])
```
```{r plotOpeLogManual}
#--------------------------------
## Save as png
png(filename='plotOpeLogManual.png')

## Draw Plot: box-plot
p <- ggplot(data, aes(x=Group, y=timesPerDay, fill=Group)) + 
  geom_boxplot(outlier.shape=21,) + theme_bw() +
  geom_point(position = position_jitterdodge()) +
  labs(title="User Operational Preference", subtitle="Manual Operations per Day") + xlab("Group") + ylab("Times")

## Show the graph
print(p)

## Close device
dev.off()

#--------------------------------
```
## Overall Performance Changes (OPC)
**A one-way ANOVA was employed to check the effect of the two lighting settings on participants’ overall performance changes. **

```{r plotOverallPerformanceChanges}
#--------------------------------
## Save as png
png(filename='plotAnovaOverall.png')

## Draw Plot: box-plot
p <- ggplot(work.dfChanges, aes(x=Group, y=chgTotal, fill=Group)) + 
  geom_boxplot(outlier.shape=21,) + theme_bw() +
  labs(title="Performance Changes from Term 1 to Term 2", subtitle="Overall: Chinese plus Mathematics") + xlab("Group") + ylab("Score Changes")

## Show the graph
print(p)

## Close device
dev.off()

#--------------------------------
```
```{r anova_Overall}
# ANOVA - Overall Changes
ano.overall <- aov(chgTotal ~ Group, data=work.dfChanges)
ano.overall
summary(ano.overall)
coefficients(ano.overall)

temp.increment <- computeIncrement(work.dfChanges$chgTotal, work.dfChanges$Total.y, work.dfChanges$Group)
sprintf("Overall socre of experimental group gained %.2f%% higher increment than control group", temp.increment[3]*100)

# Effect Size with 95% CI
eta_squared(ano.overall, ci=0.95)

# Since significance is reported, TukeyHSD needs to be performed to compare the effect of different level
TukeyHSD(ano.overall)
```

## By-subject Performance Changes (PC-S)
**A one-way MANOVA was used to examine to the response of the two subjects as well as each of them **

```{r plotBySubjectPerformanceChanges}
#--------------------------------
## Save as png
png(filename='plotAnovaBySubject.png')

## Draw Plot: box-plot
p <- ggplot(work.dfChanges, aes(fill=Group)) + 
  geom_boxplot(outlier.shape=21, aes(y=chgChinese, x='Chinese', group=Group)) +
  geom_boxplot(outlier.shape=21, aes(y=chgMath, x='Mathematics', group=Group)) + 
  labs(title="Performance Changes from Term 1 to Term 2", subtitle="By Subject") + xlab("Subject") + ylab("Score Changes") +
  #scale_y_continuous(limits = quantile(work.dfChanges$chgTotal, c(0.2, 0.8))) +
  theme_bw()

## Show the graph
print(p)

## Close device
dev.off()

#--------------------------------
```

```{r manova_BySubject}
# MANOVA - By-subject Changes
mano.subject <- manova(cbind(chgChinese, chgMath) ~ Group, data=work.dfChanges)
mano.subject

# Is changes of Chinese and Math significantly different between groups
summary.manova(mano.subject)

# Check the effect on each subject
summary.aov(mano.subject)

coefficients(mano.subject)

temp.increment <- computeIncrement(work.dfChanges$chgChinese, work.dfChanges$Chinese.y, work.dfChanges$Group)
sprintf("Chinese of experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

temp.increment <- computeIncrement(work.dfChanges$chgMath, work.dfChanges$Math.y, work.dfChanges$Group)
sprintf("Math of experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

# Effect Size with 95% CI
eta_squared(mano.subject, ci=0.95)
```
## By-gender Performance Changes (PC-G)
**A two-way ANOVA was applied to examine the impact of lighting modes on the changes of overall scores of each gender. **

```{r plotByGenderPerformanceChanges}
#--------------------------------
## Save as png
png(filename='plotAnovaByGender.png')

## Draw Plot: box-plot
p <- ggplot(work.dfChanges, aes(y=chgTotal, fill=Group, group=Group)) + 
  geom_boxplot(outlier.shape=21) +
  facet_wrap(facet = ~ Gender) +
  labs(title="Performance Changes from Term 1 to Term 2", subtitle="By Gender") + xlab("Gender") + ylab("Score Changes") +
  #scale_y_continuous(limits = quantile(work.dfChanges$chgTotal, c(0.2, 0.8))) +
  theme_bw()

## Show the graph
print(p)

## Close device
dev.off()

#--------------------------------
```
```{r plotGroupGenderInteraction}
#--------------------------------
## Save as png
png(filename='plotGender_Group.png')

p <- interaction2wt(chgTotal ~ Group + Gender, data=work.dfChanges, main.in='Main effects and 2-way interactions of PC-G', responselab='')

## Show the graph
print(p)

## Close device
dev.off()
```

```{r anova_ByGender}
# ANOVA - By gender delta
ano.gender <- aov(chgTotal ~ Group * Gender, data=work.dfChanges)
ano.gender
summary(ano.gender)
coefficients(ano.gender)

work.dfChangesGirls <- work.dfChanges[work.dfChanges$Gender=='F',]
temp.increment <- computeIncrement(work.dfChangesGirls$chgTotal, work.dfChangesGirls$Total.y, work.dfChangesGirls$Group)
sprintf("Score of girls in experimental group improved %.2f%% higher than that of in control group", temp.increment[3]*100)

work.dfChangesBoys <- work.dfChanges[work.dfChanges$Gender=='M',]
temp.increment <- computeIncrement(work.dfChangesBoys$chgTotal, work.dfChangesBoys$Total.y, work.dfChangesBoys$Group)
sprintf("Score of boys in experimental group improved %.2f%% higher than that of in control group", temp.increment[3]*100)

# Effect Size with 95% CI
eta_squared(ano.gender, ci=0.95)

# Since significance is reported, TukeyHSD needs to be performed to compare the effect of different level
TukeyHSD(ano.gender)
```
## Performance Changes By-subject and By-genders (PC-SG)
**A two-way MANOVA was performed to analyze the detailed effects on different subject and different gender. **

```{r plotBySubjectGenderPerformanceChanges}
#--------------------------------
## Save as png
png(filename='plotAnovaBySubjectGender.png')

## Draw Plot: box-plot
p <- ggplot(work.dfChanges, aes(fill=Group)) + 
  geom_boxplot(outlier.shape=21, aes(y=chgChinese, x='Chinese', group=Group)) +
  geom_boxplot(outlier.shape=21, aes(y=chgMath, x='Mathematics', group=Group)) + 
  facet_wrap(facet = ~ Gender) +
  labs(title="Performance Changes from Term 1 to Term 2", subtitle="By Subject & Gender") + xlab("Subject") + ylab("Score Changes") +
  #scale_y_continuous(limits = quantile(work.dfChanges$chgTotal, c(0.2, 0.8))) +
  theme_bw()

## Show the graph
print(p)

## Close device
dev.off()

#--------------------------------
```

```{r manova_SubjectGender}
# 2-way MANOVA: Chinese, Math ~ Group + Gender
mano.gender <- manova(cbind(chgChinese, chgMath) ~ Group + Gender, data=work.dfChanges)
mano.gender

# Is changes of Chinese and Math significantly different between groups
summary.manova(mano.gender)

# Check the effect on each subject
summary.aov(mano.gender)

coefficients(mano.gender)

temp.increment <- computeIncrement(work.dfChangesGirls$chgChinese, work.dfChangesGirls$Chinese.y, work.dfChangesGirls$Group)
sprintf("Chinese of girls in experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

temp.increment <- computeIncrement(work.dfChangesGirls$chgMath, work.dfChangesGirls$Math.y, work.dfChangesGirls$Group)
sprintf("Math of girls in experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

temp.increment <- computeIncrement(work.dfChangesBoys$chgChinese, work.dfChangesBoys$Chinese.y, work.dfChangesBoys$Group)
sprintf("Chinese of boys in experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

temp.increment <- computeIncrement(work.dfChangesBoys$chgMath, work.dfChangesBoys$Math.y, work.dfChangesBoys$Group)
sprintf("Math of boys in experimental group improved %.2f%% higher than control group", temp.increment[3]*100)

# Effect Size with 95% CI
eta_squared(mano.gender, ci=0.95)
```
```{r manova_GenderGroup}
# 2-way MANOVA: Chinese, Math ~ Gender + Group
mano.genderG <- manova(cbind(chgChinese, chgMath) ~ Gender + Group, data=work.dfChanges)
mano.genderG

# Is changes of Chinese and Math significantly different between groups
summary.manova(mano.genderG)

# Check the effect on each subject
summary.aov(mano.genderG)

coefficients(mano.genderG)

# Effect Size with 95% CI
eta_squared(mano.genderG, ci=0.95)
```

